---
title: "Outlier Analysis"
author: "Domenick Braccia"
date: "8/15/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggpubr)
```

# Overview

Here, outlier features from each of the three pipelines considered - dada2, mothur and qiime - are examined. Features are labeled outliers based on 6 different error metrics: "median_error", "iqr_error", "rcov_error", "mean_error", "var_error" and "cov_error" 

### Retrieving saved data

Data comes from relative abundance error metrics section of "relative_abundance_assessment_results.Rmd" file and is saved to the ~/data/ folder. Boxplots for each error metric are drawn from which outliers will be extracted and examined.

```{r}
# loading data saved from rel_abundance_est Rmd file #
rel_abu_error <- readRDS(file = "../data/rel_abu_error.RDS")
rel_abu_error_summary <- readRDS(file = "../data/rel_abu_error_summary.RDS")
```

```{r, warning=FALSE, echo=FALSE, message=FALSE}
######################### MEDIAN ERROR ######################### 
### Error rate boxplot and outlier annotation
median_error_boxplot <- rel_abu_error %>% group_by(pipe, biosample_id, feature_id) %>%
    summarise(median_error = median(inferred_error_rate)) %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = median_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Bias Metric", color = "Pipeline")
#error_boxplot

## Annotating features as outliers based on boxplot
median_plot_dat <- ggplot_build(median_error_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_median_dat <- median_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

median_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_median_dat) %>%
    mutate(outlier_cat = if_else(median_error < ymin | median_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_mederror <- median_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "median")
outliers_mederror <- outliers_mederror[, c(5,1,2,3,4)]
#outliers_mederror


######################### IQR ERROR ############################
## Robust IQR Analysis
iqr_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = iqr_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")
#iqr_boxplot

## Annotating features as outliers based on boxplot
iqr_plot_dat <- ggplot_build(iqr_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_iqr_dat <- iqr_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

iqr_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_iqr_dat) %>%
    mutate(outlier_cat = if_else(iqr_error < ymin | iqr_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_iqrerror <- iqr_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "iqr")
outliers_iqrerror <- outliers_iqrerror[, c(5,1,2,3,4)]
#outliers_iqrerror

######################### RCOV ERROR ###########################
## Robust COV Analysis
rcov_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = rcov_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")
#rcov_boxplot

## Annotating features as outliers based on boxplot
rcov_plot_dat <- ggplot_build(rcov_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_rcov_dat <- rcov_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

rcov_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_rcov_dat) %>%
    mutate(outlier_cat = if_else(rcov_error < ymin | rcov_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_rcoverror <- rcov_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "rcov")
outliers_rcoverror <- outliers_rcoverror[, c(5,1,2,3,4)]
#outliers_rcoverror



######################### MEAN ERROR ###########################
## MEAN error rate boxplot and outlier analysis
mean_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = mean_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")
#mean_boxplot

## Annotating features as outliers based on boxplot
mean_plot_dat <- ggplot_build(mean_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_mean_dat <- mean_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

mean_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_mean_dat) %>%
    mutate(outlier_cat = if_else(mean_error < ymin | mean_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_meanerror <- mean_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "mean")
outliers_meanerror <- outliers_meanerror[, c(5,1,2,3,4)]
#outliers_meanerror


######################### VAR ERROR ############################
## VAR error rate boxplot and outlier analysis
var_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = var_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")
#var_boxplot

## Annotating features as outliers based on boxplot
var_plot_dat <- ggplot_build(var_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_var_dat <- var_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

var_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_var_dat) %>%
    mutate(outlier_cat = if_else(var_error < ymin | var_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_varerror <- var_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "var")
outliers_varerror <- outliers_varerror[, c(6,1,2,3,4,5)]
#outliers_varerror


######################### COV ERROR ############################
## COV error rate boxplot and outlier analysis
cov_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = cov_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")
#cov_boxplot

## Annotating features as outliers based on boxplot
cov_plot_dat <- ggplot_build(cov_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_cov_dat <- cov_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

cov_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_cov_dat) %>%
    mutate(outlier_cat = if_else(cov_error < ymin | cov_error > ymax,
                                 "outlier","inlier"))

# making outliers summary table
outliers_coverror <- cov_outlier_cat %>% 
    group_by(pipe, outlier_cat) %>%
    summarize("totals" = n()) %>%
    spread(outlier_cat, totals) %>%
    mutate("outlier_prop" = outlier / inlier) %>%
    add_column(metric = "cov")
outliers_coverror <- outliers_coverror[, c(6,1,2,3,4,5)]
#outliers_coverror


##### Bringing all metric tables together (will be added to later) #####
outliers <- list(median_error = outliers_mederror, iqr_error = outliers_iqrerror, 
                 rcov_error = outliers_rcoverror, mean_error = outliers_meanerror, 
                 var_error = outliers_varerror, cov_error = outliers_coverror)
outliers_cat <- list(median_cat = median_outlier_cat, iqr_cat = iqr_outlier_cat, 
                 rcov_cat = rcov_outlier_cat, mean_cat = mean_outlier_cat, 
                 var_cat = var_outlier_cat, cov_cat = cov_outlier_cat)
```

# Plotting error distributions 

Below are distributions of the feature level error metrics examined. The vertical line on each plot indicates the value at which features are labeled "outliers"

```{r metric distributions, warning=FALSE, echo=FALSE}
############################ GENERIC ############################
# Initializing 
pipelines = c("dada2", "mothur", "qiime", "unclustered")
metric_cat <- c("median_cat", "iqr_cat", "rcov_cat", 
             "mean_cat", "var_cat", "cov_cat")
metrics <- c("median_error", "iqr_error", "rcov_error", 
             "mean_error", "var_error", "cov_error")

# looping through outlier_cat variables
#plot_hists <- function(pipelines, metric_cat, metrics) {
plot_list <- list()
for (i in seq(pipelines)) {
    for (j in seq(metrics)) {
        current_features <- filter(outliers_cat[[metric_cat[j]]], pipe == pipelines[i])[metrics[j]]
        current_outliers <- filter(outliers_cat[[metric_cat[j]]], pipe == pipelines[i], outlier_cat == "outlier")[metrics[j]]
        
        #ggplot version of plots
        colnames(current_features) <- c("data")
        plot_list[[j]] <- ggplot(current_features, aes(x = data)) +
            geom_histogram(bins = 75) + 
            geom_vline(xintercept = min(current_outliers)) +
            ggtitle(pipelines[i]) + xlab(metrics[j])
        
        ## base R way
        #hist(current_features, breaks = 100, main = paste(pipelines[i], metrics[j]))
        #abline(v = min(current_outliers))
    }
    # arrange a plot for each pipeline
    aplot <- ggarrange(plotlist = plot_list, ncol = 3, nrow = 2)
    print(aplot)
}
#}
#plot_hists(pipelines = pipelines, metric_cat = metric_cat, metrics = metrics)
```

# Examining "repeat" outliers

Below, two columns are added to the outlier summary tables. 

`repeat_prop` contains the proportion of outlier features that appear as outliers in more than one sample. Overall, there tends to be very few - if any - repeat outliers for each of the three pipelines examined.

`prop_greater_2x` contains the proportion of outlier features that are greater than 2x the minumum outlier error metric value. These proportions seems to vary greatly by pipeline.

**NOTE: for the variance and covariance error metrics, a handful of features were not identified as outliers or inlers. This seems to stem from the original definition of an outlier and inlier from earlier in the analysis, since most features have the same value for `var_error` or `cov_error` (see histograms for these metrics above).**

```{r repeat offenders, warning=FALSE}
#Q: do features within a pipeline show up as outliers for more than one sample?
#Q: how many outlier features have values 2X the smallest outlier value?

##### GENERIC #####
#repeat_outliers <- function(metric_cat) {
repeat_outliers_row <- vector(mode = "numeric")
repeat_outleirs <- matrix(0, nrow = 4, ncol = 6)
prop_greater_2x_row <- vector(mode = "numeric")
prop_greater_2x <- matrix(0, nrow = 4, ncol = 6)
for (i in seq(pipelines)) {
    for (j in seq(metrics)) {
        #calculating number of outliers > 2x the minumum outlier value
        current_outliers <- filter(outliers_cat[[metric_cat[j]]], pipe == pipelines[i], outlier_cat == "outlier")[metrics[j]]
        prop_greater_2x_row[j] <- sum(current_outliers > 2 * min(current_outliers)) / nrow(current_outliers)
        
        #making repeat outlier rows for summary tables
        current_metric <- outliers_cat[[metric_cat[j]]] %>% 
            filter(pipe == pipelines[i], outlier_cat == "outlier") %>%
            group_by(feature_id) %>%
            summarise(num_samples = n())
        repeat_outliers_row[j] <- sum(current_metric$num_samples > 1) / length(current_metric$num_samples)
    }
    #print(repeat_outliers_row)
    repeat_outleirs[i, ] <- repeat_outliers_row
    #print(prop_greater_2x_row)
    prop_greater_2x[i, ] <- prop_greater_2x_row
}

for (i in seq(metrics)) {
    outliers[[i]] %<>% 
        add_column(repeat_prop = repeat_outleirs[, i]) %>%
        add_column(prop_greater_2x = prop_greater_2x[, i])
}
print(outliers, width = Inf)
```


