---
title: "Microbiome Scale Qualitative Assessment Results"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2:
    toc: FALSE
---


```{r qualSetup, include=FALSE}
library(here)
library(tidyverse)
library(broom)
library(ggpubr)

raw_counts <-here("data","raw_counts.RDS") %>% readRDS()
nb_counts <-here("data","nb_counts_titrations.RDS") %>% readRDS()
pa_summary_anno_df <-here("data","pa_summary_anno_df.RDS") %>% readRDS()
theta_est <-here("data","bootstrap_theta_estimates.rds") %>% readRDS()
mix_binom_tidy <-here("data","mix_bayes_test.RDS") %>% readRDS()
unmix_binom_tidy <-here("data","unmix_binom_test.RDS") %>% readRDS()
```

```{r qualMunge, message = FALSE, echo = FALSE, warning = FALSE}
## Feature-PCR rep level tidy data frame
## Relevant raw_count factors
count_df <- raw_counts %>%
      dplyr::select(biosample_id, t_fctr, id, pipe, feature_id, count)

count_df <- count_df %>%
      group_by(id) %>%
      mutate(total_count = sum(count))

## Sanity Check
# count_df$total_count %>% unique() %>% summary()

## Annotating with feature type
count_df <- pa_summary_anno_df %>%
      dplyr::select(biosample_id, pipe, feature_id, pa_specific) %>%
      left_join(count_df)

## Inferred Feature Proportions
### Same code used in Rel_Abu_Error
pre_post_prop <- nb_counts %>%
      ungroup() %>%
      filter(t_fctr %in% c(0,20)) %>%
      mutate(end_point = if_else(t_fctr == 0 , "post", "pre")) %>%
      dplyr::select(-t_fctr) %>%
      ## setting values to 0 when one or more of the PCR replicates are 0 for titration end-points
      spread(end_point,nb_prop, fill = 0)

prop_inferred <- theta_est %>%
      filter(pipe == "unclustered") %>%
      ungroup() %>%
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>%
      dplyr::select(biosample_id, theta_hat_mean, t_fctr) %>%
      right_join(nb_counts) %>% right_join(pre_post_prop) %>%
      filter(t_fctr %in% c(1:5,10,15)) %>%
      ## Setting negative inferrred theta values to 0
      mutate(theta_hat_mean = if_else(theta_hat_mean < 0, 0, theta_hat_mean)) %>%
      ## Using inferred theta estimates to calculate expected values
      mutate(inferred_prop = post * theta_hat_mean + pre * (1 - theta_hat_mean))

count_df <- prop_inferred %>%
      dplyr::select(biosample_id, t_fctr, pipe, feature_id,
             pre, post, inferred_prop, theta_hat_mean) %>%
      left_join(count_df)

unmix_count_df <- filter(count_df,
                    t_fctr %in% c(1:5, 10, 15),
                    pa_specific == "unmixed",
                    # should not have to filter by inferred_prop unmix should
                    # have non-zero inferred prop or at least non-zero pre +
                    # post
                    pre + post != 0, count == 0)

mix_count_df <- filter(count_df,
                  t_fctr %in% c(1:5, 10, 15),
                  count != 0, pre == 0, post == 0)

## Multipannel plots
unmix_plot_df <- unmix_count_df %>%
      filter(pipe != "unclustered") %>%
      mutate(exp_count = inferred_prop * total_count) %>%
      ungroup() %>%
      mutate(pipe = forcats::fct_recode(pipe, DADA2 = "dada2", Mothur = "mothur", QIIME = "qiime"))

unmix_count_plot <- ggplot(unmix_plot_df) +
      geom_boxplot(aes(x = biosample_id, y = exp_count + 1),
                   alpha = 0.25) +
      geom_hline(aes(yintercept = 1), linetype = 2, color = "darkorange") +
      scale_y_log10(breaks = c(1,10,100,1000)) +
      theme_bw() +
      labs(x = "Individual", y = "Expected Counts") +
      ggtitle("Unmixed-Specific") + 
      theme(legend.position = "bottom", 
            axis.text.x = element_text(angle = -45, hjust = 0),
            plot.background = element_rect(fill = "transparent",colour = NA),
            plot.title = element_text(face = "italic")) +
      facet_wrap(~pipe, nrow = 1)

## Only including p < 0.05
unmix_binom_prop <- unmix_binom_tidy %>%
      filter(pipe != "unclustered") %>%
      mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>%
      group_by(pipe, biosample_id) %>%
      mutate(count = n()) %>%
      group_by(pipe, biosample_id, count, pval_cat) %>%
      summarise(cat_count = n()) %>%
      mutate(pval_prop = cat_count/count) %>%
      filter(pval_cat == "p < 0.05")

unmix_test_plot <- unmix_binom_prop %>%
      ungroup() %>%
      mutate(pipe = forcats::fct_recode(pipe, DADA2 = "dada2", Mothur = "mothur", QIIME = "qiime")) %>%
      ggplot() +
      geom_bar(aes(x = biosample_id, y = pval_prop), stat = "identity") +
      facet_wrap(~pipe, nrow = 1) +
      labs(x = "Individual", y = "Artifactual Feature Proportion") +
      ggtitle("Unmixed-Specific") +
      theme_bw() + 
      theme(axis.text.x = element_text(angle = -45, hjust = 0),
            plot.background = element_rect(fill = "transparent",colour = NA),
            plot.title = element_text(face = "italic")) +
      ylim(0, 1)

mix_count_plot <- mix_count_df %>% filter(pipe != "unclustered") %>%
      ungroup() %>%
      mutate(pipe = forcats::fct_recode(pipe, DADA2 = "dada2", Mothur = "mothur", QIIME = "qiime")) %>%
      ggplot() +
      geom_boxplot(aes(x = biosample_id, y = count),
                   alpha = 0.25) +
      geom_hline(aes(yintercept = 1), linetype = 2, color = "darkorange") +
      scale_y_log10(breaks = c(1,10,100,1000)) +
      theme_bw() +
      labs(x = "Individual", y = "Observed Counts")  +
      ggtitle("Titration-Specific") +
      theme(legend.position = "bottom",
            axis.text.x = element_text(angle = -45, hjust = 0),
            plot.background = element_rect(fill = "transparent",colour = NA),
            plot.title = element_text(face = "italic")) +
      facet_wrap(~pipe, nrow = 1)

## Only including p < 0.05
mix_binom_prop <- mix_binom_tidy %>%
      filter(pipe != "unclustered") %>%
      mutate(pval_cat = if_else(adj.pvalue < 0.05, "p < 0.05", "p >= 0.05")) %>%
      group_by(pipe, biosample_id) %>%
      mutate(count = n()) %>%
      group_by(pipe, biosample_id, count, pval_cat) %>%
      summarise(cat_count = n()) %>%
      mutate(pval_prop = cat_count/count) %>%
      filter(pval_cat == "p < 0.05")

mix_test_plot <- mix_binom_prop %>%
      ungroup() %>%
      mutate(pipe = forcats::fct_recode(pipe, DADA2 = "dada2", Mothur = "mothur", QIIME = "qiime")) %>%
      ggplot() +
      geom_bar(aes(x = biosample_id, y = pval_prop), stat = "identity") +
      facet_wrap(~pipe, nrow = 1) +
      labs(x = "Individual", y = "Artifactual Feature Proportion")  +
      ggtitle("Titration-Specific") +
      theme_bw() +  theme(axis.text.x = element_text(angle = -45, hjust = 0),
                        plot.title = element_text(face = "italic")) +
      ylim(0, 1)
```

```{r qualPlot, message = FALSE, echo = FALSE, warning = FALSE, fig.width = 6.68, fig.cap = "Distribution of (A) observed count values for titration-specific features and (B) expected count values for unmixed-specific features by pipeline and individual. The orange horizontal dashed line indicates a count value of 1. (C) Proportion of unmix-specific features and (D) titration-specific features with an adjusted p-value < 0.05 for the Bayesian hypothesis test and binomial test respectively. We failed to accept the null hypothesis when the p-value < 0.05, indicating that the discrepancy between the feature only being observed in the titrations or unmixed samples cannot be explained by sampling alone."}
ggarrange(mix_count_plot + rremove("x.text"), unmix_count_plot + rremove("x.text"),
          mix_test_plot, unmix_test_plot,
          labels = "AUTO",
          ncol = 2, nrow = 2, heights = c(4, 5))
```

```{r echo = FALSE, message = FALSE}
ggsave(here("figs","qualPlot-1.pdf"), width = 170, units = "mm")
```


## Artifactual Features Table
Calculating abundance of "sparse" features or titration specific features (supplementary table 3)

<!-- \begin{table} -->
<!-- \caption{\label{tab:TAandUSsummary}Summary of abundance of titration-specific and unmixed-specific features per pipeline. These numbers are only for E01JH0004 and are directly related to the proportions described in Supplemental Fig. \ref{fig:bayesPrior} for alpha = beta = 1.} -->
<!-- \centering -->
<!-- \resizebox{\linewidth}{!}{ -->
<!-- \begin{tabular}[t]{llllll} -->
<!-- \toprule -->
<!-- Pipeline & \# of features & \emph{Titration-Specific} (TS) & Artifactual TS & \emph{Unmixed-Specific} (US) & Artifactual US \\ -->
<!-- \midrule -->
<!-- DADA2    & 3144 & 725  & 471  & 880 & 167  \\ -->
<!-- Mothur   & 38358  &  8824  &  44  &  9047  & 18  \\ -->
<!-- QIIME    & 11385  & 6386  &  3  &  3726  & 75  \\ -->
<!-- \bottomrule -->
<!-- \end{tabular}} -->
<!-- \end{table} -->


```{r}
unmix_art_feature_count <- unmix_binom_tidy %>% 
  mutate(artifactul_rep = if_else(adj.pvalue < 0.05, 1, 0)) %>% 
  group_by(pipe, biosample_id, feature_id) %>% 
  summarise(n_art_reps = sum(artifactul_rep)) %>% 
  mutate(artifactual_feature = if_else(n_art_reps > 1, 1, 0)) %>% 
  group_by(pipe, biosample_id) %>% 
  summarise(n_art_features = sum(artifactual_feature))

mix_art_feature_count <- mix_binom_tidy %>% 
  mutate(artifactul_rep = if_else(adj.pvalue < 0.05, 1, 0)) %>% 
  group_by(pipe, biosample_id, feature_id) %>% 
  summarise(n_art_reps = sum(artifactul_rep)) %>% 
  mutate(artifactual_feature = if_else(n_art_reps > 1, 1, 0)) %>% 
  group_by(pipe, biosample_id) %>% 
  summarise(n_art_features = sum(artifactual_feature))

art_feature_counts <- bind_rows(unmixed = unmix_art_feature_count,
                           mixed = mix_art_feature_count,.id = "pa_specific")
```

```{r}
ts_us_specific_counts <- count_df %>% 
  select(biosample_id, pipe, feature_id, pa_specific) %>% 
  distinct() %>% 
  group_by(pipe, biosample_id, pa_specific) %>% 
  summarise(n_specific = n())
```


```{r}
art_feat_df <- art_feature_counts %>% 
  left_join(ts_us_specific_counts) %>% 
  mutate(af_prop = n_art_features/ n_specific)
```

```{r}
total_feature_df <- count_df %>% 
  select(pipe, biosample_id, feature_id) %>% 
  distinct() %>% 
  group_by(pipe, biosample_id) %>% 
  summarise(total_features = n())
```

```{r}
art_feat_df <- art_feature_counts %>% 
  left_join(ts_us_specific_counts) %>% 
  mutate(af_prop = n_art_features/ n_specific) %>% 
  left_join(total_feature_df)
```

```{r}
art_feat_tbl <- art_feat_df %>% 
  select(pipe, biosample_id, total_features, pa_specific,n_specific, n_art_features, af_prop) %>% 
  mutate(af_prop = round(af_prop, 4)) %>% 
  arrange(pipe, biosample_id, pa_specific)
```

```{r}
library(knitr)
library(kableExtra)
art_table_tex <- kable(art_feat_tbl, "latex", 
      longtable = T, booktabs = T, 
      caption = "Artifactual Feature Proportion") %>%
  kable_styling(latex_options = c("repeat_header")) %>% 
  collapse_rows(columns = 1:3, latex_hline = "major", valign = "middle")
write_lines(art_table_tex, here("results", "artifactual_table.tex"))
```

## Artifactual Feature Summary
Comparison of AF taxonomy across methods.
```{r}
make_taxa_df <- function(){
  require(phyloseq)
  ps_list <- list(
      dada2 = here("data","dada2_ps.rds"),
      mothur =  here("data","mothur_ps.rds"),
      qiime =  here("data","qiime_ps.rds")
  ) %>% 
      map(readRDS)
  
  get_taxa <- function(ps){
      taxa_df <- tax_table(ps) %>% 
          as.data.frame() %>%
          tibble::rownames_to_column(var = "feature_id") %>% 
          mutate_all(funs(str_remove(.,".__")))
      
      taxa_df <- tibble(feature_id = taxa_names(ps), 
                        f_counts = taxa_sums(ps)) %>% 
          left_join(taxa_df) 
  }
  
  ps_list %>% map_dfr(get_taxa, .id = "pipe")
}

mix_art_feature_df <- unmix_binom_tidy %>% 
  mutate(artifactul_rep = if_else(adj.pvalue < 0.05, 1, 0)) %>% 
  group_by(pipe, biosample_id, feature_id) %>% 
  summarise(n_art_reps = sum(artifactul_rep)) %>%  
  mutate(artifactual_feature = if_else(n_art_reps > 1, 1, 0)) %>% 
  left_join(make_taxa_df())
```

```{r, fig.height = 6}
mix_art_feature_df %>% 
  filter(artifactual_feature == 1) %>% 
  filter(pipe != "unclustered") %>% 
  ggplot() + 
  geom_bar(aes(x = Rank4, fill = pipe), position = "dodge") + 
  coord_flip() + 
  # facet_wrap(~pipe, nrow = 1, scales = "free_x") +
  theme_bw()
```

Clostridium and bacteriodetes are common artifactual features across methods
```{r}
clost_af_df <- mix_art_feature_df %>% 
  filter(Rank4 == "Clostridiales") %>% 
  group_by(pipe, biosample_id, Rank5) %>% 
  mutate(f_count = sum(f_counts)) %>% 
  group_by(pipe, biosample_id, Rank5, f_counts, artifactual_feature) %>% 
  summarise(count = n()) %>% 
  spread(artifactual_feature, count, fill = 0)
```

```{r}
clost_af_df %>% 
  ungroup() %>% 
  select(-f_counts, -`0`) %>% 
  filter(`1` != 0) %>% 
  group_by(pipe, biosample_id, Rank5) %>% 
  summarise(n_art = sum(`1`)) %>% 
  spread(pipe, n_art) %>% 
  arrange(Rank5)
```
A number of Ruminococcaceae and Lachnospiraceae features are artifacts for both DADA2 and QIIME across all five individuals, interestingly not artifactual for Mothur. 

```{r}
mix_art_feature_df %>% 
  filter(Rank5 == "Lachnospiraceae") %>% 
  arrange(Rank6)
```

