---
title: "Outlier Analysis"
author: "Domenick Braccia and Nate - but mostly Dom ;)"
date: '`r Sys.Date()`'
output: 
    bookdown::pdf_document2:
        df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggpubr)
```

# Overview

<!-- Here, outlier features from each of the three pipelines considered - dada2, mothur and qiime - are examined. Features are labeled outliers based on 6 different error metrics: "median_error", "iqr_error", "rcov_error", "mean_error", "var_error" and "cov_error"  -->

### Retrieving saved data

<!-- Data comes from relative abundance error metrics section of "relative_abundance_assessment_results.Rmd" file and is saved to the ~/data/ folder. Boxplots for each error metric are drawn from which outliers will be extracted and examined. -->

```{r}
# loading data saved from rel_abundance_est Rmd file #
rel_abu_error <- readRDS(file = "data/rel_abu_error.RDS")
rel_abu_error_summary <- readRDS(file = "data/rel_abu_error_summary.RDS")
```

```{r, warning=FALSE, echo=FALSE, message=FALSE}
######################### MEDIAN ERROR ######################### 
### Error rate boxplot and outlier annotation
median_error_boxplot <- rel_abu_error %>% group_by(pipe, biosample_id, feature_id) %>%
    summarise(median_error = median(inferred_error_rate)) %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = median_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Bias Metric", color = "Pipeline")

## Annotating features as outliers based on boxplot
median_plot_dat <- ggplot_build(median_error_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_median_dat <- median_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

med_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_median_dat) %>%
    mutate(outlier_cat = if_else(median_error < ymin | median_error > ymax,
                                 "outlier","inlier")) %>% 
    select(-ymin) %>% 
    rename(med_threshold = ymax,
           med_cat = outlier_cat)

######################### RCOV ERROR ###########################
## Robust COV Analysis
rcov_boxplot <- rel_abu_error_summary %>%
    ggplot() +
    geom_boxplot(aes(x = pipe, y = rcov_error, color = pipe), outlier.shape = NA) +
    facet_wrap(~biosample_id, ncol = 1) +
    theme_bw() + theme(axis.text.x = element_text(angle = 90)) +
    labs(x = "Pipeline", y = "Variance Metric", color = "Pipeline")

## Annotating features as outliers based on boxplot
rcov_plot_dat <- ggplot_build(rcov_boxplot)$data[[1]] %>%
    mutate(pipe = fct_recode(factor(group),
                             dada2 = "1",
                             mothur = "2",
                             qiime = "3",
                             unclustered = "4"),
           biosample_id = fct_recode(PANEL,
                                     E01JH0004 = "1",
                                     E01JH0011 = "2",
                                     E01JH0016 = "3",
                                     E01JH0017 = "4",
                                     E01JH0038 = "5"))
outlier_rcov_dat <- rcov_plot_dat %>%
    dplyr::select(ymin, ymax, pipe, biosample_id)

rcov_outlier_cat <- rel_abu_error_summary %>%
    left_join(outlier_rcov_dat) %>%
    mutate(outlier_cat = if_else(rcov_error < ymin | rcov_error > ymax,
                                 "outlier","inlier")) %>% 
    select(-ymin) %>% 
    rename(rcov_threshold = ymax,
           rcov_cat = outlier_cat)
```



```{r}
## Combining median error and RCOV error dataframes 
outlier_df <- rcov_outlier_cat %>% 
    left_join(med_outlier_cat)
```

```{r}
outlier_df %>% 
    group_by(pipe, biosample_id, rcov_cat) %>% 
    summarise(rcov_total = n()) %>% 
    spread(rcov_cat, rcov_total) %>% 
    mutate(rcov_prop = round(outlier/ inlier,2))
```

```{r}
outlier_df %>% 
    group_by(pipe, biosample_id, med_cat) %>% 
    summarise(med_total = n()) %>% 
    spread(med_cat, med_total) %>% 
    mutate(med_prop = round(outlier/ inlier,2))
```

Only two features were outliers for both assessment metrics.  

```{r}
outlier_df %>% 
    group_by(pipe, med_cat, rcov_cat) %>% 
    summarise(count = n()) %>% 
    spread(pipe, count, fill = 0)
```



# Plotting error distributions 

Below are distributions of the feature level error metrics examined. The vertical line on each plot indicates the value at which features are labeled "outliers"

__Median Error__
```{r}
med_outlier_cat %>% 
    ggplot() + 
    geom_histogram(aes(x = median_error), bins = 75) +
    geom_vline(aes(xintercept = med_threshold)) + 
    facet_grid(pipe ~ biosample_id, scales = "free") + 
    theme_bw()
        
```

__RCOV__  

```{r}
rcov_outlier_cat %>% 
    ggplot() + 
    geom_histogram(aes(x = rcov_error), bins = 75) +
    geom_vline(aes(xintercept = rcov_threshold)) + 
    facet_grid(pipe ~ biosample_id, scales = "free") + 
    theme_bw()
```


```{r}
rcov <- rcov_outlier_cat %>%
    mutate(value = rcov_error) %>% 
    rename(threshold = rcov_threshold,
           outlier_cat = rcov_cat)
med <- med_outlier_cat %>%
    mutate(value = median_error) %>% 
    rename(threshold = med_threshold,
           outlier_cat = med_cat)
long_outlier_df <- bind_rows(med = med, rcov = rcov, .id = "metric")

long_outlier_df %>% 
    filter(outlier_cat == "outlier") %>% 
ggplot() +
    geom_histogram(aes(x = value), bins = 75) + 
    facet_grid(pipe~metric, scales = "free") + 
    theme_bw()
```

Threshold values
```{r}
long_outlier_df %>% 
    select(metric, pipe, biosample_id, threshold) %>% 
    distinct() %>% 
    mutate(threshold = round(threshold,2)) %>% 
    spread(pipe, threshold)
```

## Relative Abundance Assessment Metric Quadrants

```{r}
rel_abu_error_summary %>%
    ggplot() +
    geom_point(aes(x = median_error, y = rcov_error, fill = log10(median_rel_abu)), pch = 21) +
    geom_hline(aes(yintercept = 2), color = "darkorange") + 
    geom_vline(aes(xintercept = 1), color = "darkorange") + 
    facet_grid(biosample_id ~ pipe, scales = "free") + 
    scale_fill_gradient(low = "white", high = "black") + 
    theme_bw() + 
    theme(legend.position = "bottom")
```

__Observations__  

- Features with large assessment metrics tend to have lower median relative abundance but not always.  
- It would be good to include some annecdotal observations about these features. 

## Supplemental
Feature relative abundance titration plot for example features from the four metric quadrants
```{r}

```


## A Brief Look At Extreme Outliers

Number of features with median error > 2 or RCOV > 5. 
These values were identified based on visualization of metric distriubitons and represent edges of the distribution.
```{r}
extreme_outliers_df <- long_outlier_df %>% 
    filter(median_error > 2 | rcov_error > 5)
nrow(extreme_outliers_df)
```

```{r}
long_outlier_df %>% 
    mutate(extreme_cat = if_else(median_error > 2 | rcov_error > 5,
                                 "extreme", outlier_cat)) %>% 
    ggplot() + 
    geom_point(aes(x = value, 
                   y = median_rel_abu, 
                   color = extreme_cat),
               pch = 21) + 
    geom_smooth(aes(x = value, 
               y = median_rel_abu), se = FALSE) + 
    facet_grid(pipe ~ metric, scales = "free") + 
    theme_bw() + 
    scale_y_log10() 
```

```{r}
extreme_outliers_df %>% 
    ggplot() + 
    geom_point(aes(x = value, 
                   y = median_rel_abu)) + 
    facet_grid(pipe ~ metric, scales = "free") + 
    theme_bw() + 
    scale_y_log10() 
```

# Session Information
## System Information
```{r}
sessioninfo::platform_info()
```


## Package Versions
```{r}
sessioninfo::package_info() %>% 
    filter(attached == TRUE) %>% 
    select(package, loadedversion, date, source) %>% 
    knitr::kable(booktabs = TRUE)
```
