---
title: "logFC Error Metrics - for edgeR 0 prior count"
author: "Nate Olson"
date: '`r Sys.Date()`'
output:
  bookdown::pdf_document2: 
    toc: FALSE
---

```{r logFC_setup, include=FALSE}
library(here)
library(tidyverse)
library(stringr)
library(broom)
```

## Objective 
Generate data frames for log fold-change assessment. 

__Data frames__  

* `logFC_pre` - log fold-change estimates and expected values for pre-specific and pre-dominant features. 
* `logFC_feature_summary` - feature-level linear model fit summary


```{r loadData}
## Excluding unclustered
### Loading Data ---------------------------------------------------------------
pa_summary_df <- readRDS(here("data","pa_summary_anno_df.RDS")) %>% 
      select(pipe, biosample_id, feature_id, 
             T00, T20, pa_mixed) 
      

logFC_df <- readRDS(here("data","logFC_edgeR_0_prior_df.rds")) %>%
      rename(feature_id = OTUname) 

## Combining logFC estimates
logFC_df <- readRDS(here("data","logFC_nb_df.rds")) %>% 
      rename(nb_logFC = logFC) %>% 
      right_join(logFC_df)

## Fixing taxonomy for consistency across pipelines
logFC_df <- logFC_df %>% 
      mutate(Rank2 = str_replace(Rank2, "p__",""),
             Rank3 = str_replace(Rank3, "c__",""),
             Rank4 = str_replace(Rank4, "o__",""),
             Rank5 = str_replace(Rank5, "f__",""),
             Rank6 = str_replace(Rank6, "g__",""),
             Rank7 = str_replace(Rank7, "s__","")) %>% 
      left_join(pa_summary_df)
```

```{r processData, include = FALSE}
### Extracting pre-specific and dominant features ------------------------------
logFC_prepost <- logFC_df %>% 
      filter(T1 == 0, T2 == 20)

logFC_pre <- logFC_prepost %>% 
      filter(logFC > 5, T20 == 4) %>% 
      group_by(biosample_id) %>% 
      rename(prepost_logFC = logFC, prepost_logCPM = logCPM,
             pre_nb = T2_nb, post_nb = T1_nb,
             prepost_nb_logFC = nb_logFC) %>% 
      ungroup() %>% 
      select(-T1, -T2, -PValue, -FDR) %>% 
      left_join(logFC_df)

### Calculate Expected logFC --------------------------------------------------- 
#Function for coverting numeric factors to their numeric values
as_numeric_fctr <- function(x) as.numeric(as.character(x))

logFC_pre <- logFC_pre %>% 
      mutate(T1_preprop = 1 - 2^-as_numeric_fctr(T1), 
             T2_preprop = 1 - 2^-as_numeric_fctr(T2),
            exp_logFC = log2(T2_preprop/T1_preprop)) %>% 
      select(-T1_preprop, -T2_preprop)

## Removing large data-frames that are no longer needed ------------------------
rm(logFC_df)
rm(pa_summary_df)
rm(logFC_prepost)
```


## log fold-change error metrics 
The slope estimate and R^2^ for feature level linear models of the expected and estimated log fold-change. The bias metric is 1-slope estimate and the model R^2^ value is the variance metric.

```{r}
## Fitting a linear model to individual features
logFC_feature_fit <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      select(pipe, biosample_id, feature_id, exp_logFC, logFC, nb_logFC) %>%
      rename(edgeR = logFC) %>% 
      gather("logFC_est","logFC", 
             -pipe, -biosample_id, -feature_id, -exp_logFC) %>%
      ## Excluding comparisons where logFC estimates are undefined for Inf 
      ## this is for negative binomial and edgeR (without prior count) logFC estimates
      filter(!is.na(logFC), logFC != Inf, logFC != -Inf) %>% 
      group_by(pipe, biosample_id, feature_id, logFC_est) %>% 
      nest() %>% 
      mutate(fit = map(data, ~lm(logFC ~ exp_logFC, data = .)))

logFC_feature_glance <- logFC_feature_fit %>% 
      mutate(fit_glance = map(fit, glance)) %>% 
      select(-data, -fit) %>% 
      unnest()

logFC_feature_tidy <- logFC_feature_fit %>% 
      mutate(fit_glance = map(fit, tidy)) %>% 
      select(-data, -fit) %>% 
      unnest()

logFC_feature_slope <- logFC_feature_tidy %>% 
      filter(term == "exp_logFC") %>% 
      select(-term) %>% 
      rename(slope = estimate)

logFC_feature_summary <- logFC_feature_slope %>% 
      select(-statistic, -p.value) %>% 
      left_join(logFC_feature_glance)
```

```{r}
logFC_feature_fit <- logFC_pre %>% 
      filter(T1 != 0, T2 != 20) %>% 
      select(pipe, biosample_id, feature_id, exp_logFC, logFC, nb_logFC) %>%
      rename(edgeR = logFC) %>% 
      gather("logFC_est","logFC", 
             -pipe, -biosample_id, -feature_id, -exp_logFC) %>%
      group_by(pipe, biosample_id, feature_id, logFC_est)
```


## Saving data frames
```{r}
saveRDS(logFC_pre, here("data","logFC_pre_0_prior.RDS"))
saveRDS(logFC_feature_summary, here("data","logFC_feature_summary_0_prior.RDS"))
```