---
title: "Qualitative Assessment (part 2)"
author: "Domenick J. Braccia"
date: `r Sys.Date()`
output:
  bookdown::pdf_document2: 
    toc: FALSE
---


# Contents

The contents of this document are as follows:

1. Unmixed-specific feature analyses
      [X] redo binomial test for unmixed-specific features w 97.5% confidence interval
     
2. Titration-specific analyses 
      [X] Replace uniform prior with beta prior and iterate
      [X] summarize results in plot (see new-figures rmd doc)

3. sparsity
      [X] Calculate sparsity at the individual level

most of what is contained in points 2 and 3 show up in supplementary material
# Loading libraries, data, and tidying 

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(ggpubr)
raw_counts <- readRDS("../data/raw_counts.RDS")
nb_counts <- readRDS("../data/nb_counts_titrations.RDS") # where does this come from?
pa_summary_anno_df <- readRDS("../data/pa_summary_anno_df.RDS")
theta_est <- readRDS("../data/bootstrap_theta_estimates.RDS")
```

```{r qual_tidy, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
## Feature-PCR rep level tidy data frame
## Relevant raw_count factors
count_df <- raw_counts %>%
      select(biosample_id, t_fctr, id, pipe, feature_id, count)

# grouping rows by `id` column and then adding column of total_counts
count_df <- count_df %>%
      group_by(id) %>%
      mutate(total_count = sum(count))

## Sanity Check
count_df$total_count %>% unique() %>% summary()

## Annotating with feature type
# essentially just reordered cols of count_df and changed what is being "grouped" by
count_df <- pa_summary_anno_df %>%
      select(biosample_id, pipe, feature_id, pa_specific) %>%
      left_join(count_df)

## Inferred Feature Proportions
### Same code used in Rel_Abu_Error
pre_post_prop <- nb_counts %>%
      ungroup() %>%
      filter(t_fctr %in% c(0,20)) %>% #this grabs all of the unmixed samples
      mutate(end_point = if_else(t_fctr == 0 , "post", "pre")) %>%
      select(-t_fctr) %>%
      ## setting values to 0 when one or more of the PCR replicates are 0 for titration end-points
      spread(end_point, nb_prop, fill = 0)

prop_inferred <- theta_est %>%
      filter(pipe == "unclustered") %>%
      ungroup() %>%
      mutate(t_fctr = factor(t_fctr, levels = c(0:5, 10, 15, 20))) %>%
      select(biosample_id, theta_hat_mean, t_fctr) %>%
      right_join(nb_counts) %>% right_join(pre_post_prop) %>%
      filter(t_fctr %in% c(1:5,10,15)) %>%
      ## Setting negative inferred theta values to 0
      mutate(theta_hat_mean = if_else(theta_hat_mean < 0, 0, theta_hat_mean)) %>%
      ## Using inferred theta estimates to calculate expected values
      mutate(inferred_prop = post * theta_hat_mean + pre * (1-theta_hat_mean))

count_df <- prop_inferred %>%
      select(biosample_id, t_fctr, pipe, feature_id,
             pre, post, inferred_prop, theta_hat_mean) %>%
      left_join(count_df)

unmix_count_df <- filter(count_df,
                    t_fctr %in% c(1:5, 10, 15),
                    pa_specific == "unmixed",
                    # should not have to filter by inferred_prop unmix should
                    # have non-zero inferred prop or at least non-zero pre +
                    # post
                    pre + post != 0, count == 0)

mix_count_df <- filter(count_df,
                  t_fctr %in% c(1:5, 10, 15),
                  count != 0, pre == 0, post == 0)
```

## 1. Unmixed-specific feature analyses

binomial testing for unmixed specific features with increased confidence interval

```{r}
unmix_binom_test <- function(unmix_count_df, output_file, confidence){
    if(file.exists(output_file)){
        return(readRDS(output_file))
    } else {
        print("could not find results file in '../data' ... running code now")
    }
    ## Binomial test - Is the true proportion of a feature within a mixture less
    ## than the inferred proportion assuming no observed counts for a titration
    ## given the number of reads sequenced for a PCR replicate.
      
    unmix_binom_test_df <- unmix_count_df %>%
        group_by(pipe, biosample_id, id, feature_id, t_fctr) %>%
        nest() %>%                                                
        mutate(binom_test = map(data, ~with(., binom.test(x = 0,  
                                                          n =  , 
                                                          p = inferred_prop, 
                                                          alternative = "less",
                                                          conf.level = confidence))),
               binom_tidy = map(binom_test, tidy))

    ## tidy-ing up results
    unmix_binom_tidy <- unmix_binom_test_df %>%
        select(pipe, biosample_id, id, feature_id, t_fctr, binom_tidy) %>%
        unnest() # error occurred on this line

    ## multiple test correction
    unmix_binom_tidy$adj.pvalue <- p.adjust(unmix_binom_tidy$p.value, method = "BH")

    ## Saving data frame as RDS to prevent having to rerun
    saveRDS(unmix_binom_tidy, output_file)
}
```

seperating by pipeline & comparing one tailed vs two tailed p-value distributions

```{r}
unmix_binom_0.95 <- unmix_binom_test(unmix_count_df, output_file = "../data/unmix_binom_test_0.95", confidence = 0.95)
unmix_binom_0.975 <- unmix_binom_test(unmix_count_df, output_file = "../data/unmix_binom_0.975.RDS", confidence = 0.975)
```

## 2. Titration specific analyses

Replacing uniform prior distribution with beta dist & vary beta parameters iteratively

```{r echo=TRUE, message = FALSE, warning=FALSE, include=FALSE}
calc_bayes_mc <- function(max_prop, total_count, obs_count = count, ALPHA, BETA){
      ## Using sapply to simulate join beta/uniform and binomal distributions
      
      # changing from simulating samples from a uniform distribution to
      # a beta distribution while varrying parameter selection on the
      # beta distribution
      
      # for pi_obs >= max_prop, switching to a beta distribution and
      # iterating over parameters alpha (shape1) and beta (shape2)
      prop_sim <- rbeta(n = 10000, shape1 = ALPHA, shape2 = BETA)
      mc_gt_count <- sapply(prop_sim, rbinom, size = total_count, n = 1)
      p_gt_max <- sum(mc_gt_count >= obs_count)/length(mc_gt_count) 
      
      ## sim distribution for P(Pi <= max_prop)
      prop_sim <- runif(n = 10000, min = 0, max = max_prop)
      mc_lt_count <- sapply(prop_sim, rbinom, size = total_count, n = 1)
      p_lt_max <- sum(mc_lt_count >= obs_count)/length(mc_lt_count)
      
      ## Bayesian Hypothesis test
      (p_lt_max * 0.5)/sum((p_lt_max * 0.5), (p_gt_max * 0.5))
}

mix_binom_bayes <- function(count_df, mix_count_df, output_file, ALPHA, BETA){
      ## Minimum expected proportion
      min_prop <- count_df %>% 
            ## Only allowing for non-zero minimum proportions
            ## This potentially results in a higher minimum expected proprotion than
            ## allowing for one of the two unmixed samples to have a 0 minimum
            ## proportion.
            mutate(pre = if_else(pre == 0, 1, pre),
                   post = if_else(post == 0, 1, post)) %>% 
            group_by(pipe, biosample_id) %>% 
            ## For some samples with the minimum expected prop is 0
            mutate(min_pre = min(pre),
                   min_post = min(post)) %>% 
            select(pipe, biosample_id, theta_hat_mean, min_pre, min_post) %>% 
            unique() %>% 
            mutate(min_exp_prop = min_post*theta_hat_mean + min_pre * (1-theta_hat_mean))
      mix_count<- mix_count_df %>% 
            filter(pa_specific == "mixed") %>% 
            left_join(min_prop) %>% 
            ## excluding Unclustered decreases the number of tests from 136k to 28k
            filter(pipe != "unclustered")
      mix_binom_test <- mix_count %>% 
            ## Minimizing the number of test performed
            select(min_exp_prop, total_count, count) %>% unique() %>% 
            rowwise() %>%
            mutate(bayes_test = calc_bayes_mc(max_prop = min_exp_prop, 
                                              total_count = total_count, 
                                              obs_count = count,
                                              ALPHA, BETA))
      mix_binom_test_df <- mix_count %>% left_join(mix_binom_test)
      mix_binom_test_df$adj.pvalue <- p.adjust(mix_binom_test_df$bayes_test, method = "BH") 
      saveRDS(mix_binom_test_df, output_file)
      #mix_binom_test_df
}
```

Iterating MC simulations for various parameterizations of alpha and beta parameters of the beta distribution.

```{r}
# library(doParallel)
# library(foreach)
# cl <- makeCluster(2)
# registerDoParallel(cl)

# # trying to parallelize with foreach
# foreach(n = 1:length(Alpha_values)) %dopar% {
#       foreach(m = 1:length(Beta_values)) %dopar% {
#             output_file <- paste("../data/mix_bayes_test_results/mix_bayes_test", "_",
#                                  Alpha_values[n], "_", Beta_values[m], sep = "")
#             print(output_file)
#             mix_binom_bayes(count_df, mix_count_df, output_file = output_file, ALPHA = A, BETA = B)
#       }
# }


# below code may take some time to run on a local system.
Alpha_values <- c(1, 2, 10, 25)
Beta_values <- c(1, 2, 10, 25)

for (A in Alpha_values) {
      for (B in Beta_values) {
            output_file <- paste("../data/mix_bayes_res/test", "_", A, "_", B, ".RDS", sep = "")
            print(output_file)
            mix_binom_bayes(count_df, mix_count_df, output_file = output_file, ALPHA = A, BETA = B)
      }
}
```

## Calculating sparsity per individual

```{r}
tmp <- raw_counts %>% 
      group_by(biosample_id, pipe, feature_id) %>% 
      summarise(total_counts = sum(count)) 


n_dada2_feat <- tmp %>% filter(biosample_id == "E01JH0004", pipe == "dada2") %>% nrow()
n_mothur_feat <- tmp %>% filter(biosample_id == "E01JH0004", pipe == "mothur") %>% nrow()
n_qiime_feat <- tmp %>% filter(biosample_id == "E01JH0004", pipe == "qiime") %>% nrow()
n_unclustered_feat <- tmp %>% filter(biosample_id == "E01JH0004", pipe == "unclustered") %>% nrow()

tmp2 <- tmp %>% 
      mutate(pres_abs = (total_counts > 0)) %>%
      group_by(biosample_id, pipe) %>%
      summarise(pres_abs = sum(pres_abs)) %>%
      spread(pipe, pres_abs)

# calculate sparsity per pipeline and individual
tmp2$dada2 <- (1 - tmp2$dada2/n_dada2_feat)
tmp2$mothur <- (1 - tmp2$mothur/n_mothur_feat)
tmp2$qiime <- (1 - tmp2$qiime/n_qiime_feat)
tmp2$unclustered <- (1 - tmp2$unclustered/n_unclustered_feat)

# adding overall sparsity row on bottom
overall_sparsity <- data.frame(biosample_id = "Overall_Sparsity", dada2 = 0.93, mothur = 0.98, qiime = 0.94)
tmp2 <- tmp2[,-5]
tmp2[6,] <- overall_sparsity
tmp2[6,1] <- "Overall"
tmp2[,-1]<- round(tmp2[,-1], 2)
tmp2
```


