---
title: "Sample Fix"
author: "Nate Olson"
date: "6/13/2019"
output: html_document
---

## Overview
Objective: Generate new phyloseq seq and mrexp objects with corrected sample ids. 

Backgroud: A sample sheet error was observed when visualzing beta diversity across samples and titrations using PCA.
Based on sample groupings and an image of the Plate map Shen sent the sample ids were corrected.

## Loading packages
```{r}
library(tidyverse)
library(phyloseq)
```

## Fix Biosample IDs
```{r}
fix_post_samples <- function(ps){
  sam_data <- sample_data(ps) %>% 
    mutate(biosample_id_fix = case_when(
                              t_fctr == 0 & biosample_id == "E01JH0004" ~ "E01JH0017", 
                              t_fctr == 0 & biosample_id == "E01JH0011" ~ "E01JH0004", 
                              t_fctr == 0 & biosample_id == "E01JH0016" ~ "E01JH0011", 
                              t_fctr == 0 & biosample_id == "E01JH0017" ~ "E01JH0016", 
                              TRUE ~ biosample_id
    ))
  sample_data(ps)$biosample_id <- sam_data$biosample_id_fix
  
  ps
}

pipeline_dir <- "~/Projects/mgtst_pipelines"
ps_list <- list(
      dada2 = file.path(pipeline_dir, "dada2/dada_ps.rds"),
      mothur =  file.path(pipeline_dir, "mothur/mothur_ps.rds"),
      qiime =  file.path(pipeline_dir, "qiime/qiime_ps.rds"),
      unclustered =  file.path(pipeline_dir, "mothur/unclustered_ps.rds")
      ) %>% 
      map(readRDS) %>% map(fix_post_samples)
```

## Save Phyloseq Objects
```{r}
for(i in 1:4){
     saveRDS(ps_list[[i]], 
             file.path("../data", paste0(names(ps_list)[i],"_ps.rds"))) 
}

```

```{r}
mrexp_list <- ps_list %>% map(phyloseq_to_metagenomeSeq)
```

```{r}
for(i in 1:4){
     saveRDS(mrexp_list[[i]], 
             file.path("../data", paste0(names(mrexp_list)[i],"_mrexp.rds"))) 
}
```

