---
title: "Fix SRA Run Table"
author: "Nate Olson"
date: "10/4/2019"
output: html_document
---

```{r}
library(here)
library(tidyverse)
```

# Objective 
Fix metadata in 16S rRNA mixture study SRA submission metadata table to account for sample mix-up. 

# Approach
Use Bioproject SRA table `https://www.ncbi.nlm.nih.gov/Traces/study/?query_key=1&WebEnv=NCID_1_15171883_130.14.22.33_5555_1584366702_1880590830_0MetA0_S_HStore&o=acc_s%3Aa` to get current sample metadata.


# Input Files

```{r}
## SRA table downloaded from run selector table 3/16 10AM
sra_tbl <- read_csv(here("data", "sra_submission_fix",
                         "SraRunTable_20200316.txt"))


## Not sure where this is file is from 
old_sra_tbl <- read_tsv(here("docs","SraRunTableAbundanceAssessment.txt"))

## SRA submission portal metadata table downloaded 3/16 10AM
sra_submission_tbl <- read_tsv(here("data", "sra_submission_fix", 
                                    "metadata-4273012-processed-ok.tsv"))

## Biosample table downloaded from submission portal
biosample_submission_tbl <- read_tsv(here("data", "sra_submission_fix", 
                                    "biosample_submission_portal_table_20200316.tsv"))
    
## Abundance assessment manuscript supplemental metadata table
supp_tbl <- read_tsv(here("data", "sra_submission_fix",
                          "40168_2020_812_MOESM3_ESM.tsv")) 
```

# Metadata tables
```{r}
glimpse(sra_tbl)
```

```{r}
glimpse(old_sra_tbl)
```

```{r}
glimpse(biosample_submission_tbl)
```

```{r}
glimpse(sra_submission_tbl)
```

```{r}
glimpse(supp_tbl)
```

# Getting Corrected Metadata table 
Making sure the table is corrected for updating submission
```{r}
check_samplesheet <- function(sample_df, fix = TRUE){
    ## Add columns with correct biosample (b_check) and titration (t_check)
    t20_pos <- paste0("A", c(1, 2, 3, 4, 5, 7, 8, 9, 10, 11))
    t0_pos <- paste0(rep(c("B", "C", "E", "F", "G"), each = 2), 
                     rep(c(6, 12), 5))
    ## Setting titration factor as character to prevent comparison errors
    sample_df <- sample_df %>% 
        mutate(t_fctr = as.character(t_fctr),
               biosample_id = as.character(biosample_id))
    
    sam_check_df <- sample_df %>%
        mutate(t_fctr = as.character(t_fctr),
            t_check = case_when(pos %in% t20_pos ~ "20",
                                pos %in% t0_pos ~ "0",
                                TRUE ~ t_fctr
            ),
            b_check = case_when(
                pos %in% c("B6", "B12") ~ "E01JH0004",
                pos %in% c("C6", "C12") ~ "E01JH0011",
                pos %in% c("E6", "E12") ~ "E01JH0016",
                pos %in% c("F6", "F12") ~ "E01JH0017",
                pos %in% c("G6", "G12") ~ "E01JH0038",
                TRUE ~ biosample_id
            )
        )
    if (all(sam_check_df$t_fctr == sam_check_df$t_check,na.rm = TRUE) |
        all(sam_check_df$biosample_id == sam_check_df$b_check, na.rm = TRUE)) {
        if (fix) {
            msg <- str_c(
                "Titration or biosample metadata is not corrected. ",
                "Returning data frame with corrected metadata"
            )
            message(msg)
            
            corrected_sample_df <- sam_check_df %>%
                mutate(t_fctr = t_check,
                       biosample_id = b_check) %>%
                select(-t_check, -b_check)
            
            return(corrected_sample_df)
        } else {
            message(str_c(
                "Titration or biosample metadata is not correct ",
                "and has not been corrected. Returning input data frame."
            ))
            return(sample_df)
        }
    }
    
    message(str_c("Titration and biosample metadata is correct", 
            ", returning input data frame"))
    return(sample_df)
}
```

```{r}
sample_sheet <- read_tsv(here("data","mgtstMetadata.tsv")) %>% 
    check_samplesheet()
```

```{r}
sample_sheet
```

```{r}
name_fix <- sample_sheet %>% 
    mutate(seq_lab_uc = toupper(seq_lab),
           seq_run = as.character(seq_run)) %>% 
    rename(PCR_plate = pcr_16S_plate, PCR_well = pos, titration = t_fctr,
           new_host_subject_id = biosample_id) %>%
    unite(new_Sample_Name, new_host_subject_id,
                                   titration,
                                   PCR_plate,
                                   PCR_well, sep = "_", remove = FALSE) %>% 
    unite("new_Library_Name", new_Sample_Name, seq_lab_uc, seq_run, 
          sep = "_",remove = FALSE) %>% 
    select(PCR_plate, PCR_well, new_Library_Name, new_Sample_Name, 
           new_host_subject_id, seq_lab, seq_run, titration)
```

```{r}
glimpse(name_fix)
```

# Applying changes to SRA metadata table

Subsetting table to only include relevant columns - thoes for use in joining with `name_fix` and columns to update.
```{r}
sra_join_tbl <- sra_tbl %>% 
    rename(seq_lab = `seq_lab (run)`,
           seq_run = `seq_run (run)`) %>% 
    select(Run, BioProject, BioSample, Experiment,
           `Library Name`, PCR_plate, PCR_well, seq_lab, seq_run)
```

```{r}
glimpse(sra_join_tbl)
```

Joining on PCR_plate, PCR_well, seq_lab, and seq_run - these are metadata variables that were unaffected by the mix-up.
```{r}
sra_join_fix <- name_fix %>% 
    mutate(seq_run = paste0(seq_lab, seq_run)) %>% 
    left_join(sra_join_tbl)
```
Fully joined table
```{r}
nrow(sra_join_fix)
```

SRA metadata for all rows
```{r}
sra_join_fix %>% filter(is.na(Run)) %>% nrow()
```


As the library names include most of the metadata we can filter the table to only include rows where the new and current library names don't match.
```{r}
sra_entries_to_change <- sra_join_fix %>% 
    filter(`Library Name` != new_Library_Name)
```

```{r}
sra_entries_to_change %>% 
    select(`Library Name`, new_Library_Name)
```
The differences are only related to the biosample_id and therefore only the following columns in the SRA table require changing.
`host_subject_id (run)`, `host_subject_id`, `Library Name`, `Sample Name`.
Will want to verify other columns, specifically thoes related to titration.

```{r}
sra_update_tbl <- sra_entries_to_change %>% 
    select(-PCR_plate, -PCR_well, -seq_lab, -seq_run, -titration) %>% 
    left_join(sra_tbl) %>% 
    select(new_Library_Name, `Library Name`, new_Sample_Name, `Sample Name`, 
           new_host_subject_id, host_subject_id, Run, BioProject, 
           BioSample, Experiment, `SRA Study`)
```

```{r}
write_tsv(sra_update_tbl, 
          here("data", "sra_submission_fix","sra_accession_table_fix.tsv"))
```

# Updating Submission
## Biosample 
- Need to send email to update biosamplehelp@ncbi.nlm.nih.gov
- Include tables with 
    - Changes: Biosample accession, new and current title, new and current sample name, and new and current host_subject_id
    - Clean version: Biosample, new title, new sample name, new host_subject_id 


```{r}
biosample_update_tbl <- sra_update_tbl %>% 
    rename(accession = BioSample) %>% 
    left_join(biosample_submission_tbl) %>% 
    mutate(new_sample_title = str_replace(sample_title, 
                                          "E01JH00[:digit:][:digit:]", 
                                          new_host_subject_id)) %>% 
    select(accession, sample_title, new_sample_title, 
           sample_name, new_Sample_Name, 
           host_subject_id, new_host_subject_id) %>% 
    distinct()
```

```{r}
write_tsv(biosample_update_tbl, 
          here("data", "sra_submission_fix","biosample_accession_table_fix.tsv"))
```

```{r}
biosample_update_tbl %>% colnames()
```

## SRA
- Can edit metadata in submission portal
    - For 64 entries: 
    - Elements to update - Title and and library name

```{r}
sra_submission_update_tbl <- sra_update_tbl %>% 
    rename(accession = Run, study = `SRA Study`) %>% 
    left_join(sra_submission_tbl) %>% 
    mutate(new_title = str_replace(title, 
                                          "E01JH00[:digit:][:digit:]", 
                                          new_host_subject_id)) %>% 
    rename(new_Biosample.name = new_Sample_Name) %>%
    select(accession, bioproject_accession, biosample_accession,
           title, new_title, library_ID, new_Library_Name)
```


```{r}
sra_submission_update_tbl
```
```{r}
write_tsv(sra_submission_update_tbl, 
          here("data", "sra_submission_fix","sra_submission_table_fix.tsv"))
```

